---
title: "8. Conditional branching"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: textmate
    css: mystyle.css
    number_sections: true
    pandoc_args: [
      "--number-offset", 8
    ]
---

```{r,echo=FALSE}
rm(list=objects()) # start with a clean workspace
source("knitr_tweaks.R")
```


A second kind of flow control that programming languages provide is the ability to evaluate **conditional statements**. Unlike loops, which can repeat over and over again, a conditional statement only executes once, but it can switch between different possible commands depending on a condition that is specified by the programmer. The power of these commands is that they allow the program itself to make choices, and in particular, to make different choices depending on the context in which the program is run. The most prominent of example of a conditional statement is the `if` statement, and the accompanying `else` statement. 

## If

The basic format of an `if` statement in R is as follows:
```
if ( CONDITION ) {
  STATEMENT1
  STATEMENT2
  ETC 
}
```

And the execution of the statement is pretty straightforward. If the condition is true, then R will execute the statements contained in the curly braces. If the condition is false, then it does not. 

## Else

If you want to, you can extend the `if` statement to include an `else` statement as well, leading to the following syntax:


```
if ( CONDITION ) {
  STATEMENT1
  STATEMENT2
  ETC
} else {
  STATEMENT3
  STATEMENT4
  ETC
}
```
As you’d expect, the interpretation of this version is similar. If the condition is true, then the contents of the first block of code (i.e., statement1, statement2, etc) are executed; but if it is false, then the contents of the second block of code (i.e., statement3, statement4, etc) are executed instead.

## Example

To give you a feel for how you can use if and else to do something useful, the example that I’ll show you is a script (`feelings.R`) that prints out a different message depending the day of the week. Here’s the script:

```{r,eval=FALSE}
if( today == "Monday" ) {
  print( "I don’t like Mondays" )
} else {
  print( "I’m a happy little automaton" )
}
```

So let's set the value of `today` to `Monday` and call the script:

```{r}
today <- "Monday"
source("./scripts/feelings.R")
```

That's very sad. However, tomorrow should be better:

```{r}
today <- "Tuesday"
source("./scripts/feelings.R")
```

There are other ways of making conditional statements in R. In particular, the `ifelse` function and the `switch` functions can be very useful in different contexts. However, my main aim in this chapter is to briefly cover the very basics, so I’ll move on.

