---
title: "Describing Data"
output:
  html_document:
    includes:
      in_header: header.html    
    toc: true
    toc_float: true
    theme: flatly
    highlight: textmate
    css: mystyle.css
    number_sections: true
    pandoc_args: [
      "--number-offset", 99
    ]
---

```{r,echo=FALSE}
rm(list=objects()) # start with a clean workspace
source("knitr_tweaks.R")
library(tidyverse,quietly = TRUE)
```

```{css,echo=FALSE}
h1{
  line-height: 100px;
}
h2{
  line-height: 80px;
}
h3{
  line-height: 60px;
}
```


> Out of every hundred people,
> those who always know better:
> fifty-two.
> <br>
> Unsure of every step:
> almost all the rest.
> <br>
> Ready to help,
> if it doesn't take long:
> forty-nine.
> <br>
> Always good,
> because they cannot be otherwise:
> four -- well, maybe five.
> <br>
> Able to admire without envy:
> eighteen.
> <br>
> Led to error
> by youth (which passes):
> sixty, plus or minus.
> <br>
> Those not to be messed with:
> four-and-forty.
> <br>
> Living in constant fear
> of someone or something:
> seventy-seven.
> <br>
> Capable of happiness:
> twenty-some-odd at most.
> <br>
> Harmless alone,
> turning savage in crowds:
> more than half, for sure.
> <br>
> Cruel
> when forced by circumstances:
> it's better not to know,
> not even approximately.
> <br>
> Wise in hindsight:
> not many more
> than wise in foresight.
> <br>
> Getting nothing out of life except things:
> thirty
> (though I would like to be wrong).
> <br>
> Balled up in pain
> and without a flashlight in the dark:
> eighty-three, sooner or later.
> <br>
> Those who are just:
> quite a few, thirty-five.
> <br>
> But if it takes effort to understand:
> three.
> <br>
> Worthy of empathy:
> ninety-nine.
> <br>
> Mortal:
> one hundred out of one hundred -- 
> a figure that has never varied yet.<br>
> <br>
> &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;- Wislawa Szymborska, [A Word on Statistics](https://www.theatlantic.com/past/docs/unbound/poetry/antholog/szymbors/stats.htm) (translated from the Polish by Joanna Trzeciak)

In this chapter I'm going to provide a quick overview of how to perform common data analysis jobs using R. It's essentially a brief coverage of descriptive statistics and a few standard inferential statistics ($t$-tests, etc). I won't explain the logic of *how* the statistical tests work: there are lots of great stats textbooks out there for that. If you're really desperate you can check out the [Learning Statistics with R](http://www.compcogscisydney.com/learning-statistics-with-r.html) book that I wrote for teaching my intro stats class. 

When writing my first ever set of statistics lectures back in 2011, I scraped a set of data from the web that contained some basic information about every Australian Football League (AFL) game played during the years 1987 to 2010. The data is stored in the `afl` data frame and saved in the `afl24.Rdata` file. We'll use that data set throughout this chapter, so here it is:

```{r} 
load("./data/afl24.Rdata") # load data
head(afl)                  # show the first few rows
```

I'm also going to make use of a couple of different packages in this chapter, so I'll load them now:

```{r}
library(psych) # William Revelle's psychometrics package
library(lsr)   # my "learning statistics with R" package (badly in need of some updating!)
library(car)   # nicer Anova function than the base packages
```

## Descriptive statistics

### Using `summary`

Let's suppose all we want to do is get a quick overview of all the variables in the data frame. The `summary` function^[Note: like `print` and `plot`, `summary` is a generic function and does rather different things depending on what kind of input you give it] provides a very handy way of doing that, as you can see here:
```{r}
summary(afl)
```
The output here is giving a sensible summary of each of the variables. To look at any one of these in a little more detail, let's start by considering the "weekday" variable, `afl$weekday`. This is a categorical variable (i.e., a *factor* in R) and so when we summarise that variable on its own, the output is a frequency table:
```{r}
summary( afl$weekday )
```
So there were a total of `r sum(afl$weekday=="Mon")` games played on a Monday over this time period, `r sum(afl$weekday=="Tue")` games played on a Tuesday, and so on. If we feed it a *numeric* variable, like `afl$attendance` we get a set of numbers describing the overall distribution:^[Just FYI: If we summarise a logical variable using a command like `summary(afl$is.final)` we get something very much like a frequency table. For a character variable the `summary` function doesn't work very well, so you'd need to explicitly tell R to treat the variable as a factor using a command like `summary(as.factor(myvariable))`]
```{r}
summary( afl$attendance )
```
So the output here gives you the range: minimum attendance was `r (min(afl$attendance)) %f% "int"` and the maximum was `r (max(afl$attendance)) %f% "int"`. The mean attendance at an AFL game was `r (mean(afl$attendance)) %f% "int"`, whereas the median was a little smaller at `r (median(afl$attendance)) %f% "int"`. Finally, it also gives you the interquartile range, which runs from  `r (quantile(afl$attendance,.25)) %f% "int"` to `r (quantile(afl$attendance,.75)) %f% "int"`. A more detailed view could be generated using the `hist` function to draw a histogram, as discussed in the graphics chapter:
```{r, echo=FALSE}
hist(
  x = afl$attendance,
  col = "lightblue",
  xlab = "Official Attendance",
  ylab = "Frequency",
  main = "",
  axes = FALSE
)
axis(
  side = 1,
  at = seq(0, 120000, 30000),
  labels = seq(0, 120000, 30000) %f% "char"
)
axis(
  side = 2
)
```


### Using `describe`

The `psych` package contains a handy function called `describe` that provides some additional summary statistics that are commonly reported in psychology. However, the measures that it reports are only really sensible for numeric variables (e.g., what exactly would the "mean" of `afl$venue` be?) so the output looks a little weird when the data set contains non-numeric variables. So to produce a tidy output, it's handy to do something like this:

```{r}
numeric.vars <- sapply(afl, is.numeric)  # I'll explain "sapply" later!
describe( afl[,numeric.vars] )           # only "describe" the numeric vars...
```

The package also contains a `describeBy` function that allows you to report the descriptive statistics by group. So, for instance let's suppose I wanted to calculate statistics separately for home and away games (i.e., `afl$is.final == FALSE`) and finals cames (i.e, `afl$is.final == TRUE`) we could use this command:

```{r}
describeBy(
  x =  afl[,numeric.vars], 
  group = afl$is.final 
)
```
This provides handy information: the average attendance for a home and away game was `r mean(afl$attendance[afl$is.final==FALSE]) %f% "int"` whereas for a finals match the average attendance was  `r mean(afl$attendance[afl$is.final==TRUE]) %f% "int"`.


### Mean and median

Introductory stats textbooks usually suggest that we describe the **central tendency** of a variable in terms of its arithmetic *mean* (i.e., the average value), the *median* (i.e., the middle value) or sometimes the *mode* (i.e., the most frequently occurring value). The mean and median are usually applied to numeric variables, whereas the mode is generally more appropriate for categorical variables. I'll talk about the mode later.

To calculate the mean or median of a variable, R supplies the `mean` and `median` functions. It's also worth noting that you can calculate the trimmed mean (e.g., 10\% trimmed mean is the mean value after removing the highest 10\% and lowest values 10\% of the data) by specifying the `trim` argument to the `mean` function. Here's a few examples:

```{r}
mean( afl$home.score ) # mean score by the home team
mean( afl$home.score, trim = 0.1 )  # 10% trimmed mean
mean( afl$home.score, trim = 0.5 )  # 50% trimmed mean...
median( afl$home.score ) # ...  median *is* the 50% trimmed mean
```

### Standard deviation and quantiles

To calculate the **spread** of a variable we often describe the *standard deviation* or report various *quantiles* of a distribution. The `sd` function will compute standard deviations for you...

```{r}
sd( afl$home.score )  # standard deviation
var( afl$home.score ) # variance, just in case you need this!
```

For quantiles, there are a number of specific functions that calculate "special" quantiles (e.g., the interquartile range is `IQR` and the range is `range`), but for the most part I think it makes sense to work with the `quantile` function that can compute any quantiles you like. Here are some examples:

```{r}
quantile(
  x = afl$home.score,
  probs = c(.25, .75) # interquartile range
)
quantile(
    x = afl$home.score,
    probs = c(0, 1)   # the range
)
quantile(
  x = afl$home.score,
  probs = c(.5)       # the median
)
```
Conventiently, the default value of `probs` is to compute all five of those quantiles so the default behaviour is very sensible:

```{r}
quantile( afl$home.score )
```

### Missing data

Something to note is that the `mean` function (and most of the other functions used for descriptive statistics) is a little fragile when given missing data. If you want to *ignore* missing values in a data set, you have to explicitly tell these functions to do so. Most of the time the name of the argument to do this is `na.rm`. For example:

```{r}
age <- c(30, 22, 24, NA)  # eek, one person didn't give their age
mean(age)                 # this doesn't produce an answer!
mean(age, na.rm = TRUE)   # ignore the missing value & average the others
```


### Tabulating data

When dealing with categorical variables, it's convenient to construct a frequency table. The `table` function^[see also the `xtabs` function] will do this nicely for us. For instance:

```{r}
table( afl$home.team )
```

The output of this function, incidentally is an R table object, an we can calculate the sample *mode* by finding the largest value in this table. Just to make this a little prettier for us, let's define a custom function:

```{r}
modal <- function(x) {
  f <- table(x)    # the frequency table
  m <- f == max(f) # which cases are equal to the maximum frequency
  modes <- f[m]    # return all these cases
  return(modes)
}
```

Here's what happens when we apply that to the `afl$home.team` variable:

```{r}
modal( afl$home.team )
```

So Essendon played more home games than anyone else over this time frame. One reason that it can be useful to use a function like the `modal` one we just wrote is that it's robust in the face of ties:

```{r}
eyes <- c("blue","blue","brown", "brown", "green") # this has two modes!
modal(eyes) # this gives a sensible answer
```

### Cross tabulations

Sometimes it is useful to cross tabulate two categorical variables. This is very easy to do with the `table` function, simply by feeding two categorical variables as the input! Let's suppose I wanted to count the number of times each team played a *home game* on each day of the week. Here's how I would do that:

```{r}
table(afl$home.team, afl$weekday)
```

Note that you can add tables to one another. So if I wasn't interested in home games specifically, I could do this:

```{r}
home.day <- table(afl$home.team, afl$weekday)
away.day <- table(afl$away.team, afl$weekday)
print( home.day + away.day ) 
```

Finally, you can construct three way tables or even higher order tables simply by passing more variable to the `table` function, though the output rapidly becomes hard to read! 


### Aggregating by group

A common task in data analysis is to compute (say) the average value of `attendance` by year. I mentioned one way to do this earlier with the `describeBy` function, but another rather more flexible way is to use `aggregate`. Here's how it works:

```{r}
att.by.year <- aggregate(
  formula = attendance ~ year, # we want to look at "attendance" broken down by "year" 
  data = afl,                  # the data are stored in the afl data frame
  FUN = mean                   # the thing we want to compute is the mean
)
att.by.year

```

That does look like quite the upward trend, which becomes more obvious when we plot the data:
```{r}
plot(
  x = att.by.year$year,
  y = att.by.year$attendance,
  xlab = "Year",
  ylab = "Average Attendance",
  type = "h",
  lwd = 5,
  ylim = c(0,40000),
  main = ""
)
```

### Correlations

The `cor` function in R is a flexible tool for calculating correlations between numeric variables. If I want to calculate the correlation between the score of the home team and the score of the away team, for instance, all I do is pass both variables to the `cor` function:
```{r}
cor( afl$home.score, afl$away.score )
```

There's not much of a relationship between them, as this scatterplot illustrates:

```{r, echo=FALSE}
plot(
  x = afl$home.score,
  y = afl$away.score,
  xlab = "Home Team Score",
  ylab = "Away Team Score",
  main = "",
  pch = 19,
  cex = .8
)
```

By default, the `cor` function computes the Pearson correlation between variables (i.e., default `method = "pearson"), but it's straightforward to compute Spearman rank order correlations:

```{r}
cor( 
  x = afl$home.score, 
  y = afl$away.score,
  method = "spearman"
)
```

The `cor` function can compute pairwise correlations between all variables in a data frame, but it only works if all the variables are numeric. Because the `afl` data has many non-numeric variables, this doesn't work:

```{r,error=TRUE}
cor(afl)
```

However, if you recall from earlier we computed a `numeric.vars` vector that indicates which ones are numeric, this version does work:

```{r}
cor(afl[,numeric.vars])
```

As an aside, if you don't like seeing so many digits in the output, you can print out a nicer summary like this:

```{r}
correlations <- cor(afl[,numeric.vars])
print(correlations, digits = 2)
```

### Computing new variables

Finally, one thing to note is that you can make your analysis more flexible by adding new variables to your data frame. For instance, suppose I'm not actually interested in the correlation between home team and away team, but rather want to look at the correlation between the winning score and the losing score (i.e., is there such a thing as a "high scoring game"?). This is where it can be *very* handy to have a good grasp of the data manipulation tools in R. Here's how we might compute the winning and losing scores using the `apply` function:

```{r}
afl$winning.score <- apply( 
  X = afl[, c("home.score","away.score")], # the home and away scores
  MARGIN = 1,                              # "retain" the first dimension
  FUN = max                                #  max along the other dimension
)

afl$losing.score <- apply( 
  X = afl[, c("home.score","away.score")], # the home and away scores
  MARGIN = 1,                              # "retain" the first dimension
  FUN = min                                #  min along the other dimension
)

head(afl)
```

There are a lot of really nice data manipulation functions in R, particularly in the various packages that make up the `tidyverse` (see [here](https://www.tidyverse.org/) for information). In the meantime, let's correlate our new variables:

```{r}
cor(afl$winning.score, afl$losing.score)
```


So there *is* a relationship between the two, but it's only modest. Here's the scatterplot:

```{r, echo=FALSE}
plot(
  x = afl$winning.score,
  y = afl$losing.score,
  xlab = "Winning Team Score",
  ylab = "Losing Team Score",
  main = "",
  pch = 19,
  cex = .8
)
```

